Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-11 vertical-thermal-review.Rmd
small adjustments to first traits paragraph
#102

Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004]. 
The strongest driver of this variation is the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022].
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig. `r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022]. 
Yet, independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy. 
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022]. 
Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.

>>>>

Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004]. 
The primary driver of this variation is the vertical light gradient, with strong differentiation between sun and shade leaves [@casasSunShadeLeaves2011; @binLeafTraitExpression2022].
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig. `r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022]. 
Yet, independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy. 
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022]. 
Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
Yet, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], and further research is needed to fully characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
